---
author: "Murugesh Narayanappa"
date: "2026-04-24"
output: html_document
---

<h1 style="text-align:center; font-size:24px; color:blue;">
Missed Mutation Correction
</h1>

<p style="text-align:center; font-size:18px; color:blue;">
Murugesh Narayanappa: 2026-04-24
</p>

##### Appending Missed Mutations from the Base Editor Design Tool
- Purpose:
   Identify and append these missed dual-edit mutations and their categories
  to the output dataframe.
- Background: The tool predicts amino acid changes by evaluating one nucleotide at   a time.However, when a codon contains "AA" or "CC" in the first two positions
  (before the wobble base), in either strand orientation (5'→3' or 3'→5') the base   editor is likely to edit both nucleotides simultaneously:
     - Adenine Base Editor (ABE): AA → GG
     - Cytosine Base Editor (CBE): CC → TT

  Since the tool only evaluates single-nucleotide edits, it does not predict
  the amino acid outcomes resulting from these probable dual-edit events by
  the base editor.
 

```{r echo = TRUE, comment = FALSE, message=FALSE, warning=FALSE}
# loading required libraries
library(stringr)
library(data.table)
library(janitor)
library(readxl)
library(openxlsx)
library(dplyr)
library(purrr)
library(gt)

```

- Functions for appending both ABE and CBE base editor missed mutations: 

```{r echo = TRUE, comment = FALSE, message = FALSE, warning = FALSE }
# ABE missed mutations: 
abe_missed_mutations <- function(mutations, categories, gRNA) {
  
  # getting positions
  get_positions <- function(mutations, pattern) {
    matches <- str_extract_all(mutations, pattern)[[1]]
    as.integer(str_extract(matches, "\\d+"))
  }
  
 # adding mutations and categories for extracted positions
  add_mutations <- function(mutations, categories, shared_pos, prefix, suffix, cat_entry) {
    for (position in shared_pos) {
      mutations  <- str_c(mutations, prefix, position, suffix, ";")
      categories <- str_c(categories, cat_entry, ";")
    }
    return(list(mutations = mutations, categories = categories))
  }
  
  # Check for Lys\\d+Glu and Lys\\d+Arg (\\d+ is a regular expression for digits)
  shared_pos <- intersect(get_positions(mutations, "Lys\\d+Glu"),
                          get_positions(mutations, "Lys\\d+Arg"))
  result     <- add_mutations(mutations, categories, shared_pos, "Lys", "Gly", "Missense")
  mutations  <- result$mutations
  categories <- result$categories
  
  # Check for Asn\\d+Asp and Asn\\d+Ser: 
  shared_pos <- intersect(get_positions(mutations, "Asn\\d+Asp"),
                          get_positions(mutations, "Asn\\d+Ser"))
  result     <- add_mutations(mutations, categories, shared_pos, "Asn", "Gly", "Missense")
  mutations  <- result$mutations
  categories <- result$categories
  
  # Check for Phe\\d+Ser and Phe\\d+Leu
  shared_pos <- intersect(get_positions(mutations, "Phe\\d+Ser"),
                          get_positions(mutations, "Phe\\d+Leu"))
  result     <- add_mutations(mutations, categories, shared_pos, "Phe", "Pro", "Missense")
  mutations  <- result$mutations
  categories <- result$categories
  
  # Check for Phe mutations and "AAA" in gRNA
  shared_pos <- get_positions(mutations, "Phe\\d+Phe")
  if (length(shared_pos) > 0 && str_detect(gRNA, "AAA")) {
    for (position in shared_pos) {
      mutations  <- str_c(mutations, "Phe", position, "Pro;")
      categories <- str_c(categories, "Missense;Missense;Missense;")
    }
  }
  
  # Check for Leu\\d+Leu and Leu\\d+Ser
  shared_pos <- intersect(get_positions(mutations, "Leu\\d+Leu"),
                          get_positions(mutations, "Leu\\d+Ser"))
  result     <- add_mutations(mutations, categories, shared_pos, "Leu", "Pro", "Missense")
  mutations  <- result$mutations
  categories <- result$categories
  
  # Deduplication of predictions
  mut_vec   <- unique(keep(str_split(mutations, ";")[[1]], nzchar))
  mutations <- str_c(str_c(mut_vec, collapse = ";"), ";")
  
  cat_vec    <- keep(str_split(categories, ";")[[1]], nzchar)
  if (length(cat_vec) > length(mut_vec)) cat_vec <- cat_vec[seq_len(length(mut_vec))]
  categories <- str_c(str_c(cat_vec, collapse = ";"), ";")
  
  return(list(mutations = mutations, categories = categories))
}
```

```{r echo = TRUE, comment = FALSE, message = FALSE, warning= FALSE}
# CBE missed mutations: 

cbe_missed_mutations <- function(mutations, categories, gRNA) {
  
  # getting positions
  get_positions <- function(mutations, pattern) {
    matches <- str_extract_all(mutations, pattern)[[1]]
    as.integer(str_extract(matches, "\\d+"))
  }
  
  # add mutations and categories for all shared positions
  add_mutations <- function(mutations, categories, shared_pos, prefix, suffix, cat_entry) {
    for (position in shared_pos) {
      mutations  <- str_c(mutations, prefix, position, suffix, ";")
      categories <- str_c(categories, cat_entry, ";")
    }
    return(list(mutations = mutations, categories = categories))
  }
  
  # Check for Val\\d+Val and Val\\d+Met
  
  shared_pos <- intersect(get_positions(mutations, "Val\\d+Val"),
                          get_positions(mutations, "Val\\d+Met"))
  result     <- add_mutations(mutations, categories, shared_pos, "Val", "Ile", "Missense")
  mutations  <- result$mutations
  categories <- result$categories
  
  # Check for Gly\\d+Asp and Gly\\d+Ser
  shared_pos <- intersect(get_positions(mutations, "Gly\\d+Asp"),
                          get_positions(mutations, "Gly\\d+Ser"))
  result     <- add_mutations(mutations, categories, shared_pos, "Gly", "Asn", "Missense")
  mutations  <- result$mutations
  categories <- result$categories
  
  # Check for Gly\\d+Glu and Gly\\d+Arg
  shared_pos <- intersect(get_positions(mutations, "Gly\\d+Glu"),
                          get_positions(mutations, "Gly\\d+Arg"))
  result     <- add_mutations(mutations, categories, shared_pos, "Gly", "Lys", "Missense")
  mutations  <- result$mutations
  categories <- result$categories
  
  # Check for Pro\\d+Ser and Pro\\d+Leu
  shared_pos <- intersect(get_positions(mutations, "Pro\\d+Ser"),
                          get_positions(mutations, "Pro\\d+Leu"))
  result     <- add_mutations(mutations, categories, shared_pos, "Pro", "Phe", "Missense")
  mutations  <- result$mutations
  categories <- result$categories
  
  # Deduplication of predictions
  mut_vec   <- unique(keep(str_split(mutations, ";")[[1]], nzchar))
  mutations <- str_c(str_c(mut_vec, collapse = ";"), ";")
  
  cat_vec    <- keep(str_split(categories, ";")[[1]], nzchar)
  if (length(cat_vec) > length(mut_vec)) cat_vec <- cat_vec[seq_len(length(mut_vec))]
  categories <- str_c(str_c(cat_vec, collapse = ";"), ";")
  
  return(list(mutations = mutations, categories = categories))
}

```

- Here is an example of how to append mutations to a dataframe:

```{r echo = TRUE, comment = FALSE, message =FALSE, warning =FALSE}
# read an excel file: 
hcm_abe_igf <- read_excel("p_and_fdr_values/hcm_abe_igf_bottom.xlsx") |> data.table()
# predicted mutation column : cleaned_mutations
# mutation_category column : mutation_category
# gRNA sequence column: sequences column

for(i in 1:nrow(hcm_abe_igf)){
  result <- abe_missed_mutations(hcm_abe_igf$cleaned_mutations[i],    hcm_abe_igf$mutation_category[i], hcm_abe_igf$gRNAs[i])
  hcm_abe_igf$cleaned_mutations[i] <- result$mutations
  hcm_abe_igf$mutation_category[i] <- result$categories
}

write.xlsx(hcm_abe_igf, "p_and_fdr_values/hcm_abe_igf_appended_mutations.xlsx")

head(hcm_abe_igf) |> gt() |> tab_options(table.font.size = 6)
```